Consensus cell types across osteosarcoma samples

### 2025-02-10 analyses/cell-type-consensus/exploratory-notebooks/03-osteosarcoma-consensus-celltypes.Rmd
clean up based on review comments and remove hot/cold

```{r}
# barchart with or without faceting
# each bar is a stacked barchart using the fill_color
# faceting is only done if a facet_variable is provided
stacked_barchart <- function(
    df, 
    fill_color, 
    facet_variable = NULL,
    colors
){

  barchart <- ggplot(df) + 
    aes(
      x = library_id, 
      y = percent_cells_annotation, 
      fill = !!sym(fill_color)
    ) +
    geom_col() + 
    scale_y_continuous(expand = c(0,0)) +
    scale_fill_manual(values = colors) +
    theme(axis.text.x = element_blank()) +
    labs(
      fill= "cell type"
    )

@@

  ggplot(immune_plot_df, aes(x = !!sym(plot_column), y = percent_immune)) +
    ggforce::geom_sina() +
    stat_summary(fun.y=median, geom="crossbar" , color = "red", linewidth = 0.2) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          plot.margin = margin(10,10,10,10)) +
    labs(
      x = "", 
      y = "Percent of cells annotated as Immune",
      title = plot_column
    )

@@

```{r}
stacked_barchart(total_order_df, fill_color = "top_celltypes", colors = all_celltype_colors)
```

@@

```{r}
stacked_barchart(total_order_df, fill_color = "top_celltypes", facet_variable = "project_id", colors = all_celltype_colors)
```

@@

```{r}
# now sort just by immune percentage
immune_pct_df <- all_results_df |> 
  dplyr::filter(immune_category == "immune") |> 
  dplyr::mutate(annotation_percent_immune = total_cells_per_annotation/total_cells_per_library)
  dplyr::group_by(library_id) |> 
  dplyr::summarize(
    percent_immune = sum(annotation_percent_immune)
  )

@@

stacked_barchart(immune_results_df, fill_color = "immune_category", colors = three_color_scheme)
```
Let's see how the percentage of immune cells compares across projects.

```{r}
stacked_barchart(immune_results_df, fill_color = "immune_category", facet_variable = "project_id", colors = three_color_scheme)
```

@@

Just for visualization purposes, let's classify these as "hot" and "cold", where "hot" tumors have an immune composition > 5%.

```{r}
immune_results_df <- immune_results_df |> 
  dplyr::mutate(
    # classify hot/cold based on immune percent > 5 %
    hot_cold = dplyr::if_else(percent_immune > .05, "hot", "cold")
  )
```

```{r}
stacked_barchart(immune_results_df, fill_color = "immune_category", facet_variable = "hot_cold", colors = three_color_scheme)
```

Our threshold is pretty arbitrary, but I think there are definitely some "cold" tumors where we see little to no immune involvement.

@@

stacked_barchart(immune_results_df, fill_color = "is_immune", colors = all_celltype_colors)
```

@@

```{r}
# do some data wrangling on the metadata columns we want to look at 
immune_results_df <- immune_results_df |> 
  dplyr::mutate(
    primary_or_metastasis = dplyr::if_else(!is.na(primary_or_metastasis), primary_or_metastasis, disease_timing),
    disease_timing_mod = dplyr::if_else(disease_timing %in% c("Initial diagnosis", "Recurrence"), disease_timing, "other"),
    tissue_location_mod = dplyr::case_when(
      stringr::str_detect(tissue_location, "Bone|femur|Humerus|Tibia") ~ "Bone",
      #stringr::str_detect(tissue_location, "femur") ~ "Bone",
      stringr::str_detect(tissue_location, "lung|Lung|Chest") ~ "lung or chest",
      .default = "other"
    )
  )
```

@@

    stacked_barchart(
      immune_results_df,
      fill_color = "immune_category",
      facet_variable = category,
      colors = three_color_scheme
    )
}) |>
  patchwork::wrap_plots(ncol = 1, guides = "collect") + 
    patchwork::plot_annotation(tag_levels = list(metadata_categories))

@@

    stacked_barchart(
      immune_results_df,
      fill_color = "top_celltypes",
      facet_variable = category,
      colors = all_celltype_colors
    )
}) |>
  patchwork::wrap_plots(ncol = 1, guides = "collect")
```

@@

```{r}
immune_plot_df <- immune_results_df |> 
  # just look at immune population across different diagnoses 
  dplyr::select(library_id, primary_or_metastasis, disease_timing_mod, tissue_location_mod, percent_immune, seq_unit) |> 
  unique()
```

```{r}
metadata_categories |> 
  purrr::map(\(category){
    sina_plot(immune_plot_df, category)
  })
```

>>>>

```{r}
# barchart with or without faceting
# each bar is a stacked barchart using the fill_color
# faceting is only done if a facet_variable is provided
stacked_barchart <- function(
    df, 
    fill_column,
    celltype_colors, # named vector where names match the values in fill_column
    facet_variable = NULL
){

  barchart <- ggplot(df) + 
    aes(
      x = library_id, 
      y = percent_cells_annotation, 
      fill = !!sym(fill_column)
    ) +
    geom_col() + 
    scale_y_continuous(expand = c(0,0)) +
    scale_fill_manual(values = celltype_colors) +
    theme(axis.text.x = element_blank()) +
    labs(
      fill= "cell type"
    )

@@

  ggplot(df, aes(x = !!sym(plot_column), y = percent_immune)) +
    ggforce::geom_sina() +
    stat_summary(fun.y=median, geom="crossbar" , color = "red", linewidth = 0.2) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          plot.margin = margin(10,10,10,10)) +
    labs(
      x = "", 
      y = "Percent of cells annotated as Immune",
      title = plot_column
    )

@@

```{r}
stacked_barchart(total_order_df, fill_column = "top_celltypes", celltype_colors = all_celltype_colors)
```

@@

```{r}
stacked_barchart(total_order_df, fill_column = "top_celltypes", facet_variable = "project_id", celltype_colors = all_celltype_colors)
```

@@

```{r}
# now sort just by immune percentage
immune_pct_df <- all_results_df |> 
  dplyr::filter(immune_category == "immune") |> 
  dplyr::mutate(annotation_percent_immune = total_cells_per_annotation/total_cells_per_library) |>
  dplyr::group_by(library_id) |> 
  dplyr::summarize(
    percent_immune = sum(annotation_percent_immune)
  )

@@

stacked_barchart(immune_results_df, fill_column = "immune_category", celltype_colors = three_color_scheme)
```
Let's see how the percentage of immune cells compares across projects.

```{r}
stacked_barchart(immune_results_df, fill_column = "immune_category", facet_variable = "project_id", celltype_colors = three_color_scheme)
```

@@

@@

stacked_barchart(immune_results_df, fill_column = "is_immune", celltype_colors = all_celltype_colors)
```

@@

```{r}
# do some data wrangling on the metadata columns we want to look at 
immune_metadata_df <- immune_results_df |> 
  dplyr::mutate(
    primary_or_metastasis = dplyr::if_else((is.na(primary_or_metastasis) &
                                             disease_timing != "Recurrence"), disease_timing, primary_or_metastasis) |>
      forcats::fct_relevel(c("Primary", "Metastasis")),
    disease_timing_mod = dplyr::if_else(disease_timing %in% c("Initial diagnosis", "Recurrence"), disease_timing, "other") |> 
      forcats::fct_relevel("other", after = Inf),
    tissue_location_mod = dplyr::case_when(
      stringr::str_detect(tissue_location, "Bone|femur|Humerus|Tibia") ~ "bone",
      #stringr::str_detect(tissue_location, "femur") ~ "Bone",
      stringr::str_detect(tissue_location, "lung|Lung|Chest") ~ "lung or chest",
      .default = "other"
    )
  ) 
```

@@

    df <- immune_metadata_df |> 
      dplyr::select(library_id, percent_cells_annotation, immune_category, {{category}}) |> 
      tidyr::drop_na() #remove any samples that are missing that metadata

    stacked_barchart(
      df,
      fill_column = "immune_category",
      facet_variable = category,
      celltype_colors = three_color_scheme
    )
}) |>
  patchwork::wrap_plots(ncol = 1, guides = "collect") + 
    patchwork::plot_annotation(tag_levels = list(metadata_categories))

@@

    df <- immune_metadata_df |> 
      dplyr::select(library_id, percent_cells_annotation, top_celltypes, {{category}}) |> 
      tidyr::drop_na() #remove any samples that are missing that metadata

    stacked_barchart(
      df,
      fill_column = "top_celltypes",
      facet_variable = category,
      celltype_colors = all_celltype_colors
    )
  }) |>
  patchwork::wrap_plots(ncol = 1, guides = "collect")
```

@@

```{r}
immune_plot_df <- immune_metadata_df |> 
  # just look at immune population across different diagnoses 
  dplyr::select(library_id, primary_or_metastasis, disease_timing_mod, tissue_location_mod, percent_immune, seq_unit) |> 
  unique()
```

```{r}
metadata_categories |> 
  purrr::map(\(category){

    df <- immune_plot_df |> 
      dplyr::select(library_id, percent_immune, {{category}}) |> 
      tidyr::drop_na() #remove any samples that are missing that metadata

    sina_plot(df, category)
  })
```
